AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction

### 2018-04-17 doc/ms.tex
Resubmitted version (about to be anyway)

{\changed In these cases,
the joint prior distribution for
the hyperparameters $\hypers$}, parameters, and data,
is written $p(\hypers, \params, \data | \info)$, and is usually factorized
in this way:
\begin{align}
p(\hypers, \params, \data | \info) &=
    p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \hypers, \info)\\
    &= p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \info)
\end{align}
where the first step is true in general by the product rule
{\changed of probability}, and the second
step assumes that knowing the parameters would make the hyperparameters
irrelevant when predicting what data would be observed. In the AMORPH model,
we introduced hyperparameters describing the typical amplitude
and width of the gaussians for the amorphous component, and
the degree to which the actual amplitudes and widths are scattered
around that typical value. This was also done for the crystalline
component.

@@

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/pgm-edited.pdf}
\caption{The PGM. The hyperparameters controlling the prior distributions
for the amorphous and crystalline gaussians are denoted collectively
by $\boldsymbol{\alpha}_{\rm amorph}$ and
$\boldsymbol{\alpha}_{\rm xtal}$ respectively.
The noise-related parameters are collected under $\boldsymbol{\sigma}$.
The data is $\boldsymbol{y}$.\label{fig:pgm-edited}}
\end{figure}

\begin{table}[!ht]
\footnotesize
\centering
\begin{tabular}{|lll|}
\hline
{\bf Quantity}      &   {\bf Meaning}   &  {\bf Prior distribution}\\
\hline
{\em Prior information}&&\\
$n$ & Number of measurements & given\\
$\{\x_1, \x_2, ..., \x_n\}$  & $\x$-values of measurements & given \\
{\changed $x_{\rm range}$} & {\changed Range of measurements} & {\changed $\max(x_1, ..., x_n) - \min(x_1, ..., x_n)$} \\
\hline
{\em Hyperparameters} & &\\
$N_{\rm xtal}$   &   Number of narrow peaks    &  $\propto 1/(N_{\rm xtal}+1)$, $N_{\rm xtal} \in \{0, 1, ..., 1000\}$ \\
$N_{\rm amorph}$   &   Number of wide peaks    &  $\propto 1/(N_{\rm amorph}+1)$,
 {\changed $N_{\rm amorph} \in\{0, 1, ..., 1000\}$} \\
{\changed $a_{\ln M, \rm xtal}$} & Typical {\changed log-mass} of narrow peaks & Laplace(0, 5)\\
{\changed $a_{\ln M, \rm amorph}$} & Typical {\changed log-mass} of wide peaks & Laplace(0, 5)\\
{\changed $b_{\ln M, \rm xtal}$} & Diversity of {\changed log-mass} of narrow peaks & Uniform(0, 5)\\
{\changed $b_{\ln M, \rm amorph}$} & Diversity of {\changed log-mass} of wide peaks & Uniform(0, 5)\\
$a_{\ln W, \rm xtal}$ & Typical log-width of narrow peaks & LogUniform$(10^{-3}x_{\rm range}, 0.05x_{\rm range})$\\
$a_{\ln W, \rm amorph}$ & Typical log-width of wide peaks & LogUniform$(0.05x_{\rm range}, x_{\rm range})$\\
$b_{\ln W, \rm xtal}$ & Diversity of log-width of narrow peaks & Uniform(0, 0.1)\\
$b_{\ln W, \rm amorph}$ & Diversity of log-width of wide peaks & Uniform(0, 0.2)\\
\hline
{\em Background parameters}&&\\
$b$       & Mean background level       & $\ln b \sim $ Laplace$(0, 5)$\\
$\{n_1, ..., n_4\}$  & Background deviations & independent Normal$(0,1)$\\
\hline
{\em Crystalline parameters}&&\\
$X_{i, \rm xtal}$ & Positions of narrow peaks &
                            Uniform$(x_{\rm min}, x_{\rm max})$ \\
$A_{i, \rm xtal}$ & {\changed Masses} of narrow peaks &
{\changed
 $\ln M_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln M, \rm xtal}, b_{\ln M, \rm xtal}\right)$} \\
$W_{i, \rm xtal}$ & Widths of narrow peaks &
 $\ln W_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln W, \rm xtal}, b_{\ln W, \rm xtal}\right)$ \\
{\changed $\nu_{\rm xtal}$} &  {\changed Shape of narrow peaks (common)} &  {\changed LogUniform$(0.1, 100)$}\\
\hline
{\em Amorphous parameters}&&\\
$X_{i, \rm amorph}$ & Positions of wide peaks &
          Uniform$(x_{\rm min}+0.3x_{\rm range}, x_{\rm max}-0.3x_{\rm range})$ \\
{\changed $M_{i, \rm amorph}$} &  {\changed Masses} of wide peaks &
 {\changed  $\ln M_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln M, \rm amorph}, b_{\ln M, \rm amorph}\right)$} \\
$W_{i, \rm amorph}$ & Widths of wide peaks &
 $\ln W_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln W, \rm amorph}, b_{\ln W, \rm amorph}\right)$ \\
\hline
{\em Noise parameters}&&\\
$\sigma_0$ &    Constant noise level  &   $\ln \sigma_0 \sim \textnormal{Laplace}(0,5)$\\
$\sigma_1$ &    Noise proportional to $\sqrt{\textnormal{model}}$   &  $\ln \sigma_1 \sim \textnormal{Laplace}(0,5)$ \\
$\nu$     &   Shape parameter for $t$-distribution&LogUniform$(1, 1000)$\\ &for noise   &   \\
\hline
{\em Data}&&\\
$\{y_1, y_2, ..., y_n\}$  &   Measurements    & independent \\ && Student-$t\left(f(x_i; \params), \sqrt{\sigma_0^2 + \sigma_1f(x_i; \params)},\nu\right)$\\
\hline
\end{tabular}
\caption{\it The prior distributions for all hyperparameters,
parameters, and the data.\label{tab:priors}}
\end{table}

@@

The AMORPH program calculates a fraction of amorphous material ($A_{\rm m}$) for each posterior sample of the parameters
using Equation~\ref{eqn:crystallinity}. Thus, measured crystallinity ($C_{\rm m}$) is easily determined from the relationship $C_{\rm m} = 1-A_{\rm m}$.
{\changed To test the applicability of AMORPH crystallinity results, we have analysed the Mars Scientific Laboratory (MSL)  CheMin instrument X-ray diffraction patterns of loose basaltic sand from the Bagnold Sand Dunes \citep[Gobabeb;][]{achilles2017, lapotre2017}
and sediments drilled from a silicic mudstone \citep[Buckskin sample;][]{morris2016}, both which have significant amorphous
materials (CheMin instrument data\footnote{\tt http://pds-geosciences.wustl.edu/msl/msl-m-chemin-4-rdr-v1/mslcmn{\textunderscore}1xxx.data.rdr4/}). CheMin diffraction patterns have been processed using the original Co K$\alpha$ ($1.7890 \dot{A}$) data as well as data transformed from a Co K$\alpha$ to Cu K$\alpha$ ($1.5406 \dot{A}$) X-ray source using Bragg's law. This recalculation simply allows for an easier visual comparison to terrestrial samples analysed with the more standard Cu K$\alpha$ X-ray source (see below). Note that for Co K$\alpha$ results, background critical points have been adjusted to 11.7--46.9\degree~$2\theta$ to maintain the linear background over the amorphous region from $\sim$ 8.8 to 2.25 $\dot{A}$ (Table ~\ref{tab:anode}). Although the results are uncalibrated, measured crystallinites provide a good first approximation of actual crystallinity, and relative differences are robust. Measured amorphous content of the Gobabeb 2 sample is 36.7 $\pm$ 2.9$\%$ (63.3$\%$ crystallinity) while the Buckskin 2 sediment contains 45.3 $\pm$ 2.9$\%$ amorphous phases (54.7 $\%$ crystallinity; Fig.~\ref{fig:Mars}). Comparatively, published FULLPAT calculations of amorphous contents are ~35 $\pm$ 15$\%$ for Gobabeb and 50 $\pm$ 15$\%$ for Buckskin \citep{achilles2017, morris2016}. }

>>>>

{\changed In these cases,
the joint prior distribution for
the hyperparameters $\hypers$}, parameters, and data,
is written $p(\hypers, \params, \data | \info)$, and is usually factorized
in this way:
\begin{align}
p(\hypers, \params, \data | \info) &=
    p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \hypers, \info)\\
    &= p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \info)
\end{align}
where the first step is true in general by the product rule
{\changed of probability}, and the second
step assumes that knowing the parameters would make the hyperparameters
irrelevant when predicting what data would be observed. In the AMORPH model,
we introduced hyperparameters describing the typical {\changed integrated area (`mass')}
and width of the peaks for the amorphous component, and
the degree to which the actual {\changed masses} and widths are scattered
around that typical value. This was also done for the crystalline
component.

@@

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/pgm-edited.pdf}
\caption{The PGM. The hyperparameters controlling the prior distributions
for the amorphous and crystalline peaks are denoted collectively
by $\boldsymbol{\alpha}_{\rm amorph}$ and
$\boldsymbol{\alpha}_{\rm xtal}$ respectively.
The noise-related parameters are collected under $\boldsymbol{\sigma}$.
The data is $\boldsymbol{y}$.\label{fig:pgm-edited}}
\end{figure}

\begin{table}[!ht]
\footnotesize
\centering
\begin{tabular}{|lll|}
\hline
{\bf Quantity}      &   {\bf Meaning}   &  {\bf Prior distribution}\\
\hline
{\em Prior information}&&\\
$n$ & Number of measurements & given\\
$\{\x_1, \x_2, ..., \x_n\}$  & $\x$-values of measurements & given \\
{\changed $x_{\rm range}$} & {\changed Range of measurements} & {\changed $\max(x_1, ..., x_n) - \min(x_1, ..., x_n)$} \\
\hline
{\em Hyperparameters} & &\\
$N_{\rm xtal}$   &   Number of narrow peaks    &  $\propto 1/(N_{\rm xtal}+1)$, $N_{\rm xtal} \in \{0, 1, ..., 1000\}$ \\
$N_{\rm amorph}$   &   Number of wide peaks    &  $\propto 1/(N_{\rm amorph}+1)$,
 {\changed $N_{\rm amorph} \in\{0, 1, ..., 1000\}$} \\
{\changed $a_{\ln M, \rm xtal}$} & Typical {\changed log-mass} of narrow peaks & Laplace(0, 5)\\
{\changed $a_{\ln M, \rm amorph}$} & Typical {\changed log-mass} of wide peaks & Laplace(0, 5)\\
{\changed $b_{\ln M, \rm xtal}$} & Diversity of {\changed log-mass} of narrow peaks & Uniform(0, 5)\\
{\changed $b_{\ln M, \rm amorph}$} & Diversity of {\changed log-mass} of wide peaks & Uniform(0, 5)\\
$a_{\ln W, \rm xtal}$ & Typical log-width of narrow peaks & LogUniform$(10^{-3}x_{\rm range}, 0.05x_{\rm range})$\\
$a_{\ln W, \rm amorph}$ & Typical log-width of wide peaks & LogUniform$(0.05x_{\rm range}, x_{\rm range})$\\
$b_{\ln W, \rm xtal}$ & Diversity of log-width of narrow peaks & Uniform(0, 0.1)\\
$b_{\ln W, \rm amorph}$ & Diversity of log-width of wide peaks & Uniform(0, 0.2)\\
\hline
{\em Background parameters}&&\\
$b$       & Mean background level       & $\ln b \sim $ Laplace$(0, 5)$\\
$\{n_1, ..., n_4\}$  & Background deviations & independent Normal$(0,1)$\\
\hline
{\em Crystalline parameters}&&\\
$X_{i, \rm xtal}$ & Positions of narrow peaks &
                            Uniform$(x_{\rm min}, x_{\rm max})$ \\
$M_{i, \rm xtal}$ & {\changed Masses} of narrow peaks &
{\changed
 $\ln M_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln M, \rm xtal}, b_{\ln M, \rm xtal}\right)$} \\
$W_{i, \rm xtal}$ & Widths of narrow peaks &
 $\ln W_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln W, \rm xtal}, b_{\ln W, \rm xtal}\right)$ \\
{\changed $\nu_{\rm xtal}$} &  {\changed Shape of narrow peaks (common)} &  {\changed LogUniform$(0.1, 100)$}\\
\hline
{\em Amorphous parameters}&&\\
$X_{i, \rm amorph}$ & Positions of wide peaks &
          Uniform$(x_{\rm min}+0.3x_{\rm range}, x_{\rm max}-0.3x_{\rm range})$ \\
{\changed $M_{i, \rm amorph}$} &  {\changed Masses} of wide peaks &
 {\changed  $\ln M_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln M, \rm amorph}, b_{\ln M, \rm amorph}\right)$} \\
$W_{i, \rm amorph}$ & Widths of wide peaks &
 $\ln W_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln W, \rm amorph}, b_{\ln W, \rm amorph}\right)$ \\
\hline
{\em Noise parameters}&&\\
$\sigma_0$ &    Constant noise level  &   $\ln \sigma_0 \sim \textnormal{Laplace}(0,5)$\\
$\sigma_1$ &    Noise proportional to $\sqrt{\textnormal{model}}$   &  $\ln \sigma_1 \sim \textnormal{Laplace}(0,5)$ \\
$\nu$     &   Shape parameter for $t$-distribution&LogUniform$(1, 1000)$\\ &for noise   &   \\
\hline
{\em Data}&&\\
$\{y_1, y_2, ..., y_n\}$  &   Measurements    & independent \\ && Student-$t\left(f(x_i; \params), \sqrt{\sigma_0^2 + \sigma_1f(x_i; \params)},\nu\right)$\\
\hline
\end{tabular}
\caption{\it The prior distributions for all hyperparameters,
parameters, and the data.\label{tab:priors}}
\end{table}

@@

The AMORPH program calculates a fraction of amorphous material ($A_{\rm m}$) for each posterior sample of the parameters
using Equation~\ref{eqn:crystallinity}. Thus, measured crystallinity ($C_{\rm m}$) is easily determined from the relationship $C_{\rm m} = 1-A_{\rm m}$.
{\changed To test the applicability of AMORPH crystallinity results, we have analysed the Mars Scientific Laboratory (MSL)  CheMin instrument X-ray diffraction patterns of loose basaltic sand from the Bagnold Sand Dunes \citep[Gobabeb;][]{achilles2017, lapotre2017}
and sediments drilled from a silicic mudstone \citep[Buckskin sample;][]{morris2016}, both which have significant amorphous
materials (CheMin instrument data\footnote{\tt http://pds-geosciences.wustl.edu/msl/msl-m-chemin-4-rdr-v1/mslcmn{\textunderscore}1xxx.data.rdr4/}). CheMin diffraction patterns have been processed using the original Co K$\alpha$ ($1.7890$ \AA) data as well as data transformed from a Co K$\alpha$ to Cu K$\alpha$ ($1.5406$ \AA) X-ray source using Bragg's law. This recalculation simply allows for an easier visual comparison to terrestrial samples analysed with the more standard Cu K$\alpha$ X-ray source (see below). Note that for Co K$\alpha$ results, background critical points have been adjusted to 11.7--46.9\degree~$2\theta$ to maintain the linear background over the amorphous region from $\sim$ 8.8 to 2.25 \AA (Table ~\ref{tab:anode}). Although the results are uncalibrated, measured crystallinites provide a good first approximation of actual crystallinity, and relative differences are robust. Measured amorphous content of the Gobabeb 2 sample is 36.7 $\pm$ 2.9$\%$ (63.3$\%$ crystallinity) while the Buckskin 2 sediment contains 45.3 $\pm$ 2.9$\%$ amorphous phases (54.7 $\%$ crystallinity; Fig.~\ref{fig:Mars}). Comparatively, published FULLPAT calculations of amorphous contents are ~35 $\pm$ 15$\%$ for Gobabeb and 50 $\pm$ 15$\%$ for Buckskin \citep{achilles2017, morris2016}. }
